---
title: "Olmypic History Trend Analytics"
author: "Group 9 - Thomas George Thomas, Yang Liu, Pratyush Pothuneedi"
date: "10/28/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## required packages

```{r}
library(tidyverse)
library(reshape2)
library(dplyr)
library(knitr)

## Uncomment below to set the working directory.
##setwd("C:/Users/Docs")
```

# 1. Data Acquisition

## importing the datasets
```{r} 
# Data of the athelets and countries
atheletes_df <- read.csv('athlete_events.csv', header = TRUE, sep = ',')
head(atheletes_df, 5) # structure of the dataset
```

```{r}
# Importing Data of the regions tied with the NOC code
regions_df <- read.csv('noc_regions.csv', header= TRUE, sep =',')
head(regions_df, 5)
```

# 2. Data Wrangling

## 2.1 Data Discovery

### Summary Statistics

```{r}
summary(atheletes_df)

```
## We can see that there are NA's in the numerical fields of Age, Height, Weight which we will be handling

```{r}
summary(regions_df)
```
### Discovering Discrete Data


```{r}
kable(
  atheletes_df %>%
    summarise(total_records=n()),
  caption = "Total Records for Atheletes"
  
)
```



```{r}
kable(
  regions_df %>%
    summarise(total_records=n()),
  caption = "Total Records in Regions"
  
)
```

Looking for Na's in all the columns


```{r}
colSums(is.na(atheletes_df))

```

We know that Medal's have NA, taking a closer look


```{r}
kable(
  atheletes_df %>%
    group_by(Medal) %>%
    summarise(total_records=n())
    ,caption="Records by Medal Count"
  
)
```
## We can see that NA values also exist for Medals which happens to be categorical data and we need to handle this in the cleaning part

## 2.2 Structuring

```{r}
head(atheletes_df,5)
```

## We can see that we need to do additional restructing as columns like "Games" which would require splitting is already split and available as Year and Season

## 2.3 Cleaning
