---
title: "Project 1: Olmypic History Trend Analytics"
author: "Group 9 - Thomas George Thomas, Yang Liu, Pratyush Pothuneedi"
date: "10/28/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
#Importing Required packages
library(tidyverse)
library(reshape2)
library(dplyr)
library(knitr)

## Uncomment below to set the working directory.
##setwd("C:/Users/Docs")
```

# 1. Data Acquisition

## Importing the datasets
```{r} 
# Data of the athelets and countries
atheletes_df <- read.csv('athlete_events.csv', header = TRUE, sep = ',')
head(atheletes_df, 5) # structure of the dataset
```

```{r}
# Importing Data of the regions tied with the NOC code
regions_df <- read.csv('noc_regions.csv', header= TRUE, sep =',')
head(regions_df, 5)
```

# 2. Data Wrangling

## 2.1 Data Discovery

### A. Summary Statistics

```{r}
summary(atheletes_df)
```
**We can see that there are NA's in the numerical fields of Age, Height, Weight which we will handle**

```{r}
summary(regions_df)
```
### B. Discovering Discrete Data


```{r}
kable(
  atheletes_df %>%
    summarise(total_records=n()),
  caption = "Total Records for Athletes Dataframe"
  
)
```



```{r}
kable(
  regions_df %>%
    summarise(total_records=n()),
  caption = "Total Records in Regions Dataframe"
  
)
```

Looking for NA's in all the columns


```{r}
# Store the cols with missing values 
list_na <- colnames(atheletes_df)[apply(atheletes_df, 2, anyNA)]
list_na
```

We have NA's for numerical data: Age, Height & Weight and for categorical data: Medal.


```{r}
kable(
  atheletes_df %>%
    group_by(Medal) %>%
    summarise(total_records=n())
    ,caption="Records by Medal Count"
  
)
```
**There are 231333 NA's for Medals which is categorical data and we need to handle this in the cleaning part**

```{r}
#looking for NA's in regions_df
kable(
  regions_df %>%
    filter(is.na(region)) %>%
    group_by(NOC,region,notes) %>%
  summarise(Total_records=n()),
  caption="Records grouped by categories"
)
```

No NA's in region_df



## 2.2 Structuring

```{r}
head(atheletes_df,5)
```

**We can see that we don't need to do additional restructuring as columns like "Games" is already split and available as Year and Season**

## 2.3 Cleaning

### Handling Missing Data

We can't filter out the NA values since the columns that exhibit them are required for our analysis. We will be filling the NA values for numerical columns like Age, Height, Weight with the **median** values since we require whole numbers. The Medals are filled with **'None'** which would signify that the athletes simply didn't win any of the categories of Medals (Gold, Silver, Bronze).

**We didn't filter out the NA records in Age, Height and Weight because that would mean that crucial data would be dropped leading to data skewness, we are using the Median values since we require whole numbers and to reduce the degree of skewness while maintaining data integrity.**

#### Replacing NA's in Medals

```{r, results=FALSE}
atheletes_df$Medal <- atheletes_df$Medal %>% 
  replace_na("None") # It is assumed tha the athlete participated in the sport but didn't win a medal
```

Calculating Missing Median for the missing values for Age, Height and Weight

```{r}

list_na <- list_na[ list_na !="Medal"]

# Calculate median for the missing values
missing_median <- apply(atheletes_df[,colnames(atheletes_df) %in% list_na],
      2, # 2 is for Columns
      median,
      na.rm =  TRUE)
missing_median
```


```{r}
# Replace the missing values with median
atheletes_df <- atheletes_df %>%
   mutate(
     Age  = ifelse(is.na(Age), missing_median[1], Age),
     Height = ifelse(is.na(Height), missing_median[2], Height),
     Weight = ifelse(is.na(Weight), missing_median[3], Weight)
   )
```

#### Handling Missing data in Regions


```{r}
# Replacing Na's with the respective region/notes for the NOC's
regions_df$region <- ifelse(is.na(regions_df$region), regions_df$notes, regions_df$region)
```


```{r}
kable(
  regions_df %>%
    filter(is.na(region)) %>%
    group_by(region) %>%
    summarise(total_records=n())
  ,caption = "Number of NA's in Region after fix"
  
)

```



## 2.4 Enriching

### A. Adding Attribute region

We will join regions_df and atheletes_df based on the NOC code to get the Region for enriching the data.


```{r}
athletes <- left_join(atheletes_df, regions_df, by="NOC")

# Replacing Region with Country to make the data more meaningful
colnames(athletes)[which(names(athletes) == "region")] <- "Region" 
# Removing notes since it's not relevant to our analysis anymore
athletes <- athletes[,-17]
head(athletes,5)

```

**We don't have any other attribute to split or to create a new category since we believe that we have all the required columns for our analysis**

## 2.5 Validating

### Check for any missing values

```{r}

 # Counting the number of NA's for all the columns
 colnames(athletes)[apply(athletes, 2, anyNA)]

kable(
  athletes %>%
    select(NOC,Region)  %>%
    filter(is.na(Region)) %>%
    group_by(NOC,Region) %>%
    summarise(total_records=n())
    ,caption="Null Records check by Medal Count"
  
)

```
**For NOC SGP, there are no records in our regions_df but is present in atheltes_df, as a result we are getting NA values after the join. We will add Singapore Region to the NOC in the joined data** 

```{r}

athletes$Region <- ifelse((is.na(athletes$Region) & athletes$NOC=='SGP'), "Singapore", athletes$Region)


kable(
  athletes %>%
    select(Region)  %>%
    filter(is.na(Region)) %>%
    group_by(Region) %>%
    summarise(total_records=n())
    ,caption="Checking for NA records in Region after change"
  
)

```




# Check for Duplicates

```{r}
sum(duplicated(athletes))
```
**There 1385 duplicate records on the whole data set**


```{r}
# Removing the duplicates
athletes <- unique(athletes)

```




# Checking boundary cases

```{r}
kable (
athletes %>%
summarise(max_age=max(Age), min_age=min(Age), Average_Age=mean(Age)),
caption="Age boundary cases"

)

kable (
athletes %>%
summarise(max_height=max(Height), min_height=min(Height), Average_height=mean(Height)),
caption="Height boundary cases"

)

kable (
athletes %>%
summarise(max_weight=max(Weight), min_weigt=min(Weight), Average_weight=mean(Weight)),
caption="Weight boundary cases"

)
```
All our boundary cases looks reasonable and accurate.

# 2.6 Publishing 
The data is cleaned & wrangled and made available for the team to develop business cases.


# 3. Analytical Questions

### 1. Trend analyis of Top 10 regions with the highest number of medals between 1896 - 1956 & 1957 - 2016
 
```{r fig.align="center", echo = FALSE,fig.width = 10}

p1 <-
athletes %>%
    filter(Medal!='None' & Year<=1956) %>%
    group_by(Region) %>%
    summarize(total_medals=n()) %>%
    arrange(desc(total_medals))  %>%
    mutate(Region=factor(Region, levels=Region)) %>%
    slice(1:10) %>%
    ggplot( aes(x=Region, y=total_medals))+
    geom_col(fill="steelblue") +  
    theme_minimal()+
    labs(y="Number of Medals")+
    xlab("")+
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle("Top 10 Regions with highest Medals from 1896 - 1956")+
    geom_vline(xintercept = 0)+
    geom_hline(yintercept = 0)

p2 <-
athletes %>%
    filter(Medal!='None' & Year>1956) %>%
    group_by(Region) %>%
    summarize(total_medals=n()) %>%
    arrange(desc(total_medals))  %>%
    mutate(Region=factor(Region, levels=Region)) %>%
    slice(1:10) %>%
    ggplot( aes(x=Region, y=total_medals))+
    geom_col(fill="steelblue") +  
    theme_minimal()+
    labs(y="Number of Medals")+
    xlab("")+
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle("Top 10 Regions with highest Medals from 1957 - 2016")+
    geom_vline(xintercept = 0)+
    geom_hline(yintercept = 0)

grid.arrange(p1, p2, ncol=1)


```

**Observation**: USA remains the region with the highest number of Medals in the combined history of 120 years in Olympics. Russia, a new inclusion in the top 10 took 2nd position in the later half.Germany moved into the 3rd position in the second half while UK and France slipped from 2nd and 3rd to 8th and 9th position. There are new countries in the later half such as Australia, Canada and Japan which where not in the top 10 for the earlier history of Olympics. 

###  2. Medals won by Males/Females over Time

```{r fig.width=10}

athletes %>% 
  filter(Medal!='None') %>%
  group_by(Year, Sex) %>%
  summarize(Total_Medals = n()) %>%
  ggplot(., aes(x=Year, y=Total_Medals, group=Sex, color=Sex))+
  geom_point(size=2) +
  geom_line()  +
  scale_color_manual(values=c("darkblue","red")) +
  labs(title = "Medals won by Male/Female over Time") +
  theme(plot.title = element_text(hjust = 0.5))
```

**Observation:** From the graph, we can see that there is a gradual increase in the number of medals won by female athletes over time. Male athletes tend to outnumber female athletes but their numbers also keep fluctuating over time. After the years 1994, the summer and winter Olympic games were split and held during seperate years, hence why the graph shows different points.   

##  3. Finding the most participated sport in Olympics every year

```{r}

q3<-
athletes %>% 
  group_by(Year, Sport) %>%
  summarize(Participation = n()) %>%
  arrange(Year,desc(Participation))
  
  q3 <- q3[!duplicated(q3$Year),] # Taking only the first record
  
  kable(
  q3,
  caption="Most participated Sport in Olympic Games every year"
  )

```
**Observation**: This table shows that Athletics has remained the most contested sport in 120 years of Olympics. Art Competitions were the highest participated Olympic Sport in 1932 before it was removed from the Olympics. As the Olympic Winter and Summer games were seperated into different years from 1994, Cros Country Skiing emerged as the most participated game held during the Winters.



#### 7. Top 10 host cities with the highest participation.

```{r}

##Top 10 host cities with highest participation.
Top10 <-
  athletes %>%
    select(Year,City) %>%
    group_by(Year,City) %>%
    summarize(number=n())  %>%
    arrange(desc(number))

kable(
  Top10 
  ,caption=("Top 10 host cities with highest participation")
)

##observation:
```

**Observation:**  The top ten cities with the highest participation are Sydney(2000), Atlanta(1996), Rio de Janeiro(2016),Beijing(2008), Athena(2004), Barcelona(1992), London(2012),Seoul(1988),Munich(1972),Los Angeles(1984)

#### 8. Top 10 athletes with the highest participation in the Olympics

```{r fig.width=15}
##
Top10a <-
    athletes %>%
    select(Name) %>%
    group_by(Name) %>%
    summarize(number=n())  %>%
    arrange(desc(number)) %>%
    slice(0:10)

ggplot(Top10a, aes(x=Name, y=number))+
geom_col(fill="gold") +
theme_minimal()+
labs(y="Number of Olympics ")+
xlab("athlete")+
theme(plot.title = element_text(hjust = 0.5)) +
ggtitle("Top 10 most participated athlete in the Olympics")+
geom_vline(xintercept = 0)+
geom_hline(yintercept = 0)


```

**Observation:** As we can see nearly all the top 10 athletes participated in the Olympics more than 30 times and the one with the most is Heikki Ilmari Savolainen	with 39 times

#### 9. Athlete with the most number of medals per sport

```{r}
###Athlete with the most number of medals per sport
df <- 
   athletes%>%
   filter( Medal != 'None') %>%
   select(Name,Sport)%>%
   group_by(Name,Sport) %>%
   summarize(number=n())  %>%
   arrange(desc(number)) 

dfuevent <- df[!duplicated(df$Sport),]
 
kable(
  dfuevent
  ,caption=("Athlete with the most number of medals per sport")
  )

```

**Observation:** For the Swimming, Michael Fred Phelps, II won the most number of medals(28) and there are 11 players with the number of medals equal or more than 10 in different sports.


