---
title: "Project 1: Olmypic History Trend Analytics"
author: "Group 9 - Thomas George Thomas, Yang Liu, Pratyush Pothuneedi"
date: "10/28/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE}
#Importing Required packages
library(tidyverse)
library(reshape2)
library(dplyr)
library(knitr)
library(gridExtra)
library(ggplot2)
library(data.table)
```

# 1. Introduction

## Data Description

We are considering 120 years of Olympic history where we find some interesting trends after analysis. There are two files in our data set:

1.The file *athlete_events.csv* contains 271116 rows and 15 columns. Each row corresponds to an individual athlete competing in an individual Olympic event (athlete-events). The columns are:

- ID : Unique number for each athlete
- Name : Athlete's name
- Sex : M or F
- Age : Integer
- Height : In centimeters
- Weight : In kilograms
- Team : Team name
- NOC : National Olympic Committee 3-letter code
- Games : Year and season
- Year : Integer
- Season : Summer or Winter
- City : Host city
- Sport : Sport
- Event : Event
- Medal : Gold, Silver, Bronze, or NA

2.The file *noc_regions.csv* contains 230 rows and 3 columns. Each row shows the special NOC code that denotes a region/country along with notes. The columns are:

- NOC : National Olympic Committee 3 letter code
- Country name : matches with regions in map_data("world")
- Notes : Special notes if any

We take a lot at 9 business questions that we want answered.

# Data Acquisition

We acquire the data set from Kaggle: https://www.kaggle.com/heesoo37/120-years-of-olympic-history-athletes-and-results?select=athlete_events.csv


```{r, echo=FALSE} 
# Data of the athelets and countries
atheletes_df <- read.csv('athlete_events.csv', header = TRUE, sep = ',')
```

Looking at the structure of Data set 1 - Athletes Data. Previewing only a few columns for PDF formatting

```{r, echo=FALSE}
# Structure of the dataset 1 - Athletes

kable(
  atheletes_df %>%
  select(Name,Sex,Age,Height,Weight,Medal,Sport) %>%
  slice(0:5)
)
```

```{r, echo=FALSE}
# Importing Data of the regions tied with the NOC code
regions_df <- read.csv('noc_regions.csv', header= TRUE, sep =',')
```

Looking at the structure of Data set 2 - Region Data

```{r, echo=FALSE}
# Structure of the dataset 2 - Region Data
kable(
head(regions_df, 5)
)
```


# 2. Data Wrangling

## 2.1 Data Discovery

### A. Summary Statistics

Computing summary of the athletes data (Data set 1)

```{r, echo=FALSE}
summary(atheletes_df)
```
We can see that there are NA's in the numerical fields of Age, Height, Weight which we will handle

Computing summary of region data (Data set 2)

```{r, echo=FALSE}
summary(regions_df)
```
### B. Discovering Discrete Data


```{r, echo=FALSE}
kable(
  atheletes_df %>%
    summarise(total_records=n()),
  caption = "Total Records for Athletes Dataframe"
  
)
```



```{r, echo=FALSE}
kable(
  regions_df %>%
    summarise(total_records=n()),
  caption = "Total Records in Regions Dataframe"
  
)
```

Looking for NA's in all the columns


```{r, echo=FALSE}
# Store the cols with missing values 
list_na <- colnames(atheletes_df)[apply(atheletes_df, 2, anyNA)]
list_na
```

We have NA's for numerical data: Age, Height & Weight and for categorical data: Medal.


```{r, echo=FALSE}
kable(
  atheletes_df %>%
    group_by(Medal) %>%
    summarise(total_records=n())
    ,caption="Records by Medal Count"
  
)
```
There are 231333 NA's for Medals which is categorical data and we need to handle this in the cleaning part

```{r, echo=FALSE, message=FALSE}
#looking for NA's in regions_df
kable(
  regions_df %>%
    filter(is.na(region)) %>%
    group_by(NOC,region,notes) %>%
  summarise(Total_records=n()),
  caption="Records grouped by categories"
)
```

There are 3 records with NA in region_df



## 2.2 Structuring

We can see that we don't need to do additional restructuring as columns like "Games" is already split and available as Year and Season

## 2.3 Cleaning

### A. Handling Missing Data

We don't filter out the NA values since the columns that exhibit them are required for our analysis and we don't want to drop crucial data which would lead to data skewness. We will be filling the NA values for numerical columns like Age, Height, Weight with the **median** values since we require whole numbers and to reduce the degree of skewness while maintaining data integrity. The Medals are filled with **'None'** which would signify that the athletes simply didn't win any of the categories of Medals (Gold, Silver, Bronze).

#### B. Replacing NA's in Medals

```{r, echo=FALSE}
atheletes_df$Medal <- atheletes_df$Medal %>% 
  replace_na("None") # It is assumed that the athlete participated in the sport but didn't win a medal
```

Calculating Missing Median for the missing values for Age, Height and Weight

```{r, echo=FALSE}

list_na <- list_na[ list_na !="Medal"]

# Calculate median for the missing values
missing_median <- apply(atheletes_df[,colnames(atheletes_df) %in% list_na],
      2, # 2 is for Columns
      median,
      na.rm =  TRUE)
missing_median
```


```{r, echo=FALSE}
# Replace the missing values with median
atheletes_df <- atheletes_df %>%
   mutate(
     Age  = ifelse(is.na(Age), missing_median[1], Age),
     Height = ifelse(is.na(Height), missing_median[2], Height),
     Weight = ifelse(is.na(Weight), missing_median[3], Weight)
   )
```

#### C. Handling Missing data in Regions

We replace the Na's values with region/notes for respective NOC's for the region data.
```{r, echo=FALSE}
# Replacing Na's with the respective region/notes for the NOC's
regions_df$region <- ifelse(is.na(regions_df$region), regions_df$notes, regions_df$region)
```


```{r, echo=FALSE}
kable(
  regions_df %>%
    filter(is.na(region)) %>%
    group_by(region) %>%
    summarise(total_records=n())
  ,caption = "Number of NA's in Region after fix"
)

```



## 2.4 Enriching

### A. Adding Attribute region

We will join regions_df and atheletes_df based on the NOC code to get the Region for enriching the data.


```{r, echo=FALSE, results='hide'}
athletes <- left_join(atheletes_df, regions_df, by="NOC")

# Replacing Region with Country to make the data more meaningful
colnames(athletes)[which(names(athletes) == "region")] <- "Region" 
# Removing notes since it's not relevant to our analysis anymore
athletes <- athletes[,-17]
head(athletes,5)

```

We don't have any other attribute to split or to create a new category since we believe that we have all the required columns for our analysis.

## 2.5 Validating

### A. Check for any missing values

```{r, echo=FALSE, message=FALSE}

 # Counting the number of NA's for all the columns
 colnames(athletes)[apply(athletes, 2, anyNA)]

kable(
  athletes %>%
    select(NOC,Region)  %>%
    filter(is.na(Region)) %>%
    group_by(NOC,Region) %>%
    summarise(total_records=n())
    ,caption="Null Records check by Medal Count"
  
)

```
For NOC SGP, there are no records in our regions_df but is present in atheltes_df, as a result we are getting NA values after the join. We will add Singapore Region to the NOC in the joined data. Applying fix:

```{r, echo=FALSE}

athletes$Region <- ifelse((is.na(athletes$Region) & athletes$NOC=='SGP'), "Singapore", athletes$Region)


kable(
  athletes %>%
    select(Region)  %>%
    filter(is.na(Region)) %>%
    group_by(Region) %>%
    summarise(total_records=n())
    ,caption="Checking for NA records in Region after change"
  
)

```




### B. Check for Duplicates

```{r, echo=FALSE}
sum(duplicated(athletes))
```
There **1385** duplicate records on the whole data set. Taking unique values to remove duplicates.


```{r, echo=FALSE}
# Removing the duplicates
athletes <- unique(athletes)

```


### C. Checking boundary cases

```{r, echo=FALSE}
kable (
athletes %>%
summarise(max_age=max(Age), min_age=min(Age), Average_Age=mean(Age)),
caption="Age boundary cases"

)

kable (
athletes %>%
summarise(max_height=max(Height), min_height=min(Height), Average_height=mean(Height)),
caption="Height boundary cases"

)

kable (
athletes %>%
summarise(max_weight=max(Weight), min_weigt=min(Weight), Average_weight=mean(Weight)),
caption="Weight boundary cases"

)
```
All our boundary cases looks reasonable and accurate after our wrangling.

## 2.6 Publishing 
The data is cleaned & wrangled and made available for the team to develop business cases.


# 3. Analytical Questions

### 1. Trend analysis of Top 10 regions with the highest number of medals between 1896 - 1956 & 1957 - 2016
 
```{r fig.align="center",fig.width = 10, echo=FALSE}
p1 <-
athletes %>%
    filter(Medal!='None' & Year<=1956) %>%
    group_by(Region) %>%
    summarize(total_medals=n()) %>%
    arrange(desc(total_medals))  %>%
    mutate(Region=factor(Region, levels=Region)) %>%
    slice(1:10) %>%
    ggplot( aes(x=Region, y=total_medals))+
    geom_col(fill="steelblue") +  
    theme_minimal()+
    labs(y="Number of Medals")+
    xlab("")+
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle("Top 10 Regions with highest Medals from 1896 - 1956")+
    geom_vline(xintercept = 0)+
    geom_hline(yintercept = 0)

p2 <-
athletes %>%
    filter(Medal!='None' & Year>1956) %>%
    group_by(Region) %>%
    summarize(total_medals=n()) %>%
    arrange(desc(total_medals))  %>%
    mutate(Region=factor(Region, levels=Region)) %>%
    slice(1:10) %>%
    ggplot( aes(x=Region, y=total_medals))+
    geom_col(fill="steelblue") +  
    theme_minimal()+
    labs(y="Number of Medals")+
    xlab("")+
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle("Top 10 Regions with highest Medals from 1957 - 2016")+
    geom_vline(xintercept = 0)+
    geom_hline(yintercept = 0)

grid.arrange(p1, p2, ncol=1)


```

**Observation**: USA remains the region with the highest number of Medals in the combined history of 120 years in Olympics. Russia, a new inclusion in the top 10 took 2nd position in the later half.Germany moved into the 3rd position in the second half while UK and France slipped from 2nd and 3rd to 8th and 9th position. There are new countries in the later half such as Australia, Canada and Japan which where not in the top 10 for the earlier history of Olympics. 

###  2. Medals won by Males/Females over Time

```{r fig.width=10, echo=FALSE, message=FALSE}

athletes %>% 
  filter(Medal!='None') %>%
  group_by(Year, Sex) %>%
  summarize(Total_Medals = n()) %>%
  ggplot(., aes(x=Year, y=Total_Medals, group=Sex, color=Sex))+
  geom_point(size=2) +
  geom_line()  +
  scale_color_manual(values=c("darkblue","red")) +
  labs(title = "Medals won by Male/Female over Time") +
  theme(plot.title = element_text(hjust = 0.5))
```

**Observation:** From the graph, we can see that there is a gradual increase in the number of medals won by female athletes over time. Male athletes tend to outnumber female athletes but their numbers also keep fluctuating over time. After the years 1994, the summer and winter Olympic games were split and held during seperate years, hence why the graph shows different points.   

###  3. Finding the most participated sport in Olympics every year

```{r, echo=FALSE, message=FALSE}

q3<-
athletes %>% 
  group_by(Year, Sport) %>%
  summarize(Participation = n()) %>%
  arrange(Year,desc(Participation))
  
  q3 <- q3[!duplicated(q3$Year),] # Taking only the first record
  
  kable(
  q3,
  caption="Most participated Sport in Olympic Games every year"
  )

```
**Observation**: This table shows that Athletics has remained the most contested sport in 120 years of Olympics. Art Competitions were the highest participated Olympic Sport in 1932 before it was removed from the Olympics. As the Olympic Winter and Summer games were seperated into different years from 1994, Cros Country Skiing emerged as the most participated game held during the Winters.

### 4. In which Olympic year did a particular country win a medal for the first time for a particular sport

```{r, echo=FALSE, message=FALSE}
Ans1 <-
athletes %>%
filter(Sport=="Football",Medal!="None") %>%
select(Region,Year) %>%
group_by(Region,Year) %>%
summarise(Year=min(Year))

Ans1 <- Ans1[!duplicated(Ans1$Region),]

kable (
  Ans1,
  caption="First year in which countries won medal in Football"
)

```

**Observation:** From the above table we can see the first year in which each country won a medal for football. The first countries to win medals for football are UK,Belgium,France and Netherlands and all these countries are from Europe.

### 5. Trend analysis per sport per country for the gap between medals for Football
```{r, echo=FALSE}
athletes %>%
filter(Sport=="Football",Medal!="None") %>%
select(Region, Year) %>%
group_by(Region) %>%
summarise(Number_of_Years=max(Year)-min(Year)) %>%
arrange(desc(Number_of_Years)) %>%
mutate(Region=factor(Region, levels=Region)) %>%
slice(1:20) %>%
ggplot( aes(y=Region, x=Number_of_Years))+
geom_col(fill="orange") +
theme_minimal()+
#labs(y="Region")+
xlab("Number of Years")+
theme(plot.title = element_text(hjust = 0.5)) +
ggtitle("Gap between medals for countries for Football")+
geom_vline(xintercept = 0)+
geom_hline(yintercept = 0)
```

**Observation:** The gap between first and last medals for football for each country is displayed. Canada has the largest gap of 112 years while Czech Republic has the smallest gap of 16 years.


### 6. Comparison of medals won by regions in Summer & Winter.
```{r fig.width=10, echo=FALSE}
Ans3.1 <-
athletes %>%
filter(Medal!='None',Season=="Summer") %>%
group_by(Region) %>%
summarize(number_of_medals=n()) %>%
arrange(desc(number_of_medals)) %>%
mutate(Region=factor(Region, levels=Region)) %>%
slice(1:20) %>%
ggplot( aes(y=Region, x=number_of_medals))+
geom_col(fill="green") +
theme_minimal()+
#labs(y="Region")+
xlab("Medals")+
theme(plot.title = element_text(hjust = 0.5)) +
ggtitle("Number of medals won by countries in Summer")+
geom_vline(xintercept = 0)+
geom_hline(yintercept = 0)

Ans3.2 <-
athletes %>%
filter(Medal!='None',Season=="Winter") %>%
group_by(Region) %>%
summarize(number_of_medals=n()) %>%
arrange(desc(number_of_medals)) %>%
mutate(Region=factor(Region, levels=Region)) %>%
slice(1:20) %>%
ggplot( aes(y=Region, x=number_of_medals))+
geom_col(fill="purple") +
theme_minimal()+
#labs(y="Region")+
xlab("Medals")+
theme(plot.title = element_text(hjust = 0.5)) +
ggtitle("Number of medals won by countries in Winter")+
geom_vline(xintercept = 0)+
geom_hline(yintercept = 0)

grid.arrange(Ans3.1,Ans3.2,ncol=1)

```

**Observation:** In Summer USA takes the top spot in the number of medals won and Russia stood second. When it comes to winter the positions are interchanged. Germany remains constant in both summer and winter. The graph has a uniform increase in the number of medals in summer but its not a uniform increase in winter, So this means that the winter games are more competitive while summer has a distinctive winner. 

### 7. Top 10 Host cities with the highest participation.

```{r, echo=FALSE, message=FALSE}
## Top 10 Host cities with highest participation.
Top10 <-
  athletes %>%
    select(Year,City) %>%
    group_by(Year,City) %>%
    summarize(number=n()) %>%
    arrange(desc(number))

kable(
  head(Top10, 10)
  ,caption=("Host cities with highest participation")
)


```

**Observation:**  The top ten cities with the highest participation are Sydney(2000), Atlanta(1996), Rio de Janeiro(2016),Beijing(2008), Athina(2004), Barcelona(1992), London(2012),Seoul(1988),Munich(1972),Los Angeles(1984)

### 8. Top 10 athletes with the highest participation in the Olympics

```{r fig.align="left", fig.width=15, echo=FALSE}
##
Top10a <-
    athletes %>%
    select(Name) %>%
    group_by(Name) %>%
    summarize(number=n())  %>%
    arrange(desc(number)) %>%
    slice(0:10)

ggplot(Top10a, aes(x=Name, y=number))+
geom_col(fill="gold") +
theme_minimal()+
labs(y="Number of Olympics ")+
xlab("Athletes")+
theme(plot.title = element_text(hjust = 0.5)) +
ggtitle("Top 10 most participated athlete in the Olympics")+
geom_vline(xintercept = 0)+
geom_hline(yintercept = 0)


```

**Observation:** As we can see nearly all the top 10 athletes participated in the Olympics more than 30 times and the one with the most is Heikki Ilmari Savolainen	with 39 times

### 9. Athletes with the most number of medals in each sport

```{r, echo=FALSE, message=FALSE}
###Athlete with the most number of medals per sport
df <- 
   athletes %>%
   filter( Medal != 'None') %>%
   select(Name,Sport)%>%
   group_by(Name,Sport) %>%
   summarize(number=n())  %>%
   arrange(desc(number)) 

dfuevent <- df[!duplicated(df$Sport),]
 
kable(
  head(dfuevent,11)
  ,caption=("Athlete with the most number of medals per sport")
  )

```

**Observation:** For Swimming, Michael Fred Phelps, II won the most number of medals(28) and there are 11 players with the number of medals equal or more than 10 in different sports.

# Summary

After careful analysis of Olympic history worth 120 years. We were able to decipher lot of emerging patterns and visualize them. We were able to gain valuable insights about our business questions through various plots.

**All students of Group 9 - Thomas George Thomas, Yang Liu, Pratyush Pothuneedi contributed equally to the project.**
