---
title: "Project 1: Olmypic History Trend Analytics"
author: "Group 9 - Thomas George Thomas, Yang Liu, Pratyush Pothuneedi"
date: "10/28/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
#Importing Required packages
library(tidyverse)
library(reshape2)
library(dplyr)
library(knitr)
library(gridExtra)



## Uncomment below to set the working directory.
##setwd("C:/Users/Docs")
```

# 1. Data Acquisition

## Importing the datasets
```{r} 
# Data of the athelets and countries
atheletes_df <- read.csv('athlete_events.csv', header = TRUE, sep = ',')
head(atheletes_df, 5) # structure of the dataset
```

```{r}
# Importing Data of the regions tied with the NOC code
regions_df <- read.csv('noc_regions.csv', header= TRUE, sep =',')
head(regions_df, 5)
```

# 2. Data Wrangling

## 2.1 Data Discovery

### A. Summary Statistics

```{r}
summary(atheletes_df)
```
**We can see that there are NA's in the numerical fields of Age, Height, Weight which we will handle**

```{r}
summary(regions_df)
```
### B. Discovering Discrete Data


```{r}
kable(
  atheletes_df %>%
    summarise(total_records=n()),
  caption = "Total Records for Athletes Dataframe"
  
)
```



```{r}
kable(
  regions_df %>%
    summarise(total_records=n()),
  caption = "Total Records in Regions Dataframe"
  
)
```

Looking for NA's in all the columns


```{r}
# Store the cols with missing values 
list_na <- colnames(atheletes_df)[apply(atheletes_df, 2, anyNA)]
list_na
```

We have NA's for numerical data: Age, Height & Weight and for categorical data: Medal.


```{r}
kable(
  atheletes_df %>%
    group_by(Medal) %>%
    summarise(total_records=n())
    ,caption="Records by Medal Count"
  
)
```
**There are 231333 NA's for Medals which is categorical data and we need to handle this in the cleaning part**

```{r}
#looking for NA's in regions_df
kable(
  regions_df %>%
    filter(is.na(region)) %>%
    group_by(NOC,region,notes) %>%
  summarise(Total_records=n()),
  caption="Records grouped by categories"
)
```

No NA's in region_df



## 2.2 Structuring

```{r}
head(atheletes_df,5)
```

**We can see that we don't need to do additional restructuring as columns like "Games" is already split and available as Year and Season**

## 2.3 Cleaning

### Handling Missing Data

We can't filter out the NA values since the columns that exhibit them are required for our analysis. We will be filling the NA values for numerical columns like Age, Height, Weight with the **median** values since we require whole numbers. The Medals are filled with **'None'** which would signify that the athletes simply didn't win any of the categories of Medals (Gold, Silver, Bronze).

**We didn't filter out the NA records in Age, Height and Weight because that would mean that crucial data would be dropped leading to data skewness, we are using the Median values since we require whole numbers and to reduce the degree of skewness while maintaining data integrity.**

#### Replacing NA's in Medals

```{r, results=FALSE}
atheletes_df$Medal <- atheletes_df$Medal %>% 
  replace_na("None") # It is assumed tha the athlete participated in the sport but didn't win a medal
```

Calculating Missing Median for the missing values for Age, Height and Weight

```{r}

list_na <- list_na[ list_na !="Medal"]

# Calculate median for the missing values
missing_median <- apply(atheletes_df[,colnames(atheletes_df) %in% list_na],
      2, # 2 is for Columns
      median,
      na.rm =  TRUE)
missing_median
```


```{r}
# Replace the missing values with median
atheletes_df <- atheletes_df %>%
   mutate(
     Age  = ifelse(is.na(Age), missing_median[1], Age),
     Height = ifelse(is.na(Height), missing_median[2], Height),
     Weight = ifelse(is.na(Weight), missing_median[3], Weight)
   )
```

#### Handling Missing data in Regions


```{r}
# Replacing Na's with the respective region/notes for the NOC's
regions_df$region <- ifelse(is.na(regions_df$region), regions_df$notes, regions_df$region)
```


```{r}
kable(
  regions_df %>%
    filter(is.na(region)) %>%
    summarise(total_records=n())
  ,caption = "Number of NA's in Region after fix"
  
)

```



## 2.4 Enriching

### A. Adding Attribute region

We will join regions_df and atheletes_df based on the NOC code to get the Region for enriching the data.


```{r}
athletes <- left_join(atheletes_df, regions_df, by="NOC")

# Replacing Region with Country to make the data more meaningful
colnames(athletes)[which(names(athletes) == "region")] <- "Region" 
# Removing notes since it's not relevant to our analysis anymore
athletes <- athletes[,-17]
head(athletes,5)

```

**We don't have any other attribute to split or to create a new category since we believe that we have all the required columns for our analysis**

## 2.5 Validating

### Check for any missing values

```{r}

 # Counting the number of NA's for all the columns
 colnames(athletes)[apply(athletes, 2, anyNA)]

kable(
  athletes %>%
    select(NOC,Region)  %>%
    filter(is.na(Region)) %>%
    group_by(NOC,Region) %>%
    summarise(total_records=n())
    ,caption="Null Records check by Medal Count"
  
)

```
**For NOC SGP, there are no records in our regions_df but is present in atheltes_df, as a result we are getting NA values after the join. We will add Singapore Region to the NOC in the joined data** 

```{r}

athletes$Region <- ifelse(is.na(athletes$Region) & athletes$NOC=='SGP', "Singapore", athletes$Region)


kable(
  athletes %>%
    select(Region)  %>%
    filter(is.na(Region)) %>%
    group_by(Region) %>%
    summarise(total_records=n())
    ,caption="Checking for NA records in Region after change"
  
)

```




# Check for Duplicates

```{r}
sum(duplicated(athletes))
```
**There 1385 duplicate records on the whole data set**


```{r}
# Removing the duplicates
athletes <- unique(athletes)

```




# Checking boundary cases

```{r}
kable (
athletes %>%
summarise(max_age=max(Age), min_age=min(Age), Average_Age=mean(Age)),
caption="Age boundary cases"

)

kable (
athletes %>%
summarise(max_height=max(Height), min_height=min(Height), Average_height=mean(Height)),
caption="Height boundary cases"

)

kable (
athletes %>%
summarise(max_weight=max(Weight), min_weigt=min(Weight), Average_weight=mean(Weight)),
caption="Weight boundary cases"

)
```
All our boundary cases looks reasonable and accurate.

# 2.6 Publishing 
The data is cleaned & wrangled and made available for the team to develop business cases.

# In which Olympic year did a particular country win a medal for the first time for a particular sport

```{r}
Ans1 <-
athletes %>%
filter(Sport=="Football",Medal!="None") %>%
select(Region,Year) %>%
group_by(Region,Year) %>%
summarise(Year=min(Year))

Ans1 <- Ans1[!duplicated(Ans1$Region),]

kable (
  Ans1,
  caption="First year in which countries won medal in Football"
)

```

Observation: From the above table we can see the first year in which each country won a medal for football. The first countries to win medals for football are UK,Belgium,France and Netherlands and all these countries are from Europe.

# Trend analysis per sport per country for the gap between medals for Football
```{r}

athletes %>%
filter(Sport=="Football",Medal!="None") %>%
select(Region, Year) %>%
group_by(Region) %>%
summarise(Number_of_Years=max(Year)-min(Year)) %>%
arrange(desc(Number_of_Years)) %>%
mutate(Region=factor(Region, levels=Region)) %>%
slice(1:20) %>%
ggplot( aes(y=Region, x=Number_of_Years))+
geom_col(fill="orange") +
theme_minimal()+
#labs(y="Region")+
xlab("Number of Years")+
theme(plot.title = element_text(hjust = 0.5)) +
ggtitle("Gap between medals for countries for Football")+
geom_vline(xintercept = 0)+
geom_hline(yintercept = 0)




```
Observation: The gap between first and last medals for football for each country is displayed. Canada has the largest gap while Czech Republic has the smallest gap. 

# Comparison of medals won by regions in Summer & Winter.
```{r fig.width=10}
#Summer
#df <-
# athletes %>%
# filter(Medal!="None",Season=="Summer") %>%
# select(Region) %>%
# group_by(Region) %>%
# summarise(number_of_medals=n()) %>%
# ggplot( aes(x=Region, y=number_of_medals))+
# geom_col(fill="steelblue") +
# coord_flip() +
# theme_minimal()+
# labs(y="Number of Medals")+
# xlab("")+
# theme(plot.title = element_text(hjust = 0.5)) +
# ggtitle("Total number of Medals won by countries in Summer")+
# geom_vline(xintercept = 0)+
# geom_hline(yintercept = 0)

Ans3.1 <-
athletes %>%
filter(Medal!='None',Season=="Summer") %>%
group_by(Region) %>%
summarize(number_of_medals=n()) %>%
arrange(desc(number_of_medals)) %>%
mutate(Region=factor(Region, levels=Region)) %>%
slice(1:20) %>%
ggplot( aes(y=Region, x=number_of_medals))+
geom_col(fill="green") +
theme_minimal()+
#labs(y="Region")+
xlab("Medals")+
theme(plot.title = element_text(hjust = 0.5)) +
ggtitle("Number of medals won by countries in Summer")+
geom_vline(xintercept = 0)+
geom_hline(yintercept = 0)

Ans3.2 <-
athletes %>%
filter(Medal!='None',Season=="Winter") %>%
group_by(Region) %>%
summarize(number_of_medals=n()) %>%
arrange(desc(number_of_medals)) %>%
mutate(Region=factor(Region, levels=Region)) %>%
slice(1:20) %>%
ggplot( aes(y=Region, x=number_of_medals))+
geom_col(fill="purple") +
theme_minimal()+
#labs(y="Region")+
xlab("Medals")+
theme(plot.title = element_text(hjust = 0.5)) +
ggtitle("Number of medals won by countries in Winter")+
geom_vline(xintercept = 0)+
geom_hline(yintercept = 0)

grid.arrange(Ans3.1,Ans3.2,ncol=1)

```

Observation: In Summer USA takes the top spot in the number of medals won and Russia stood second. When it comes to winter the positions are interchanged. Germany remains constant in both summer and winter. The graph has a uniform increase in the number of medals in summer but its not a uniform increase in winter, So this means that the winter games are more competitive while summer has a distinctive winner. 